```{r setup, include=FALSE}
# packages to be installed 
install_and_load_packages <- function() {
  # List of packages to be installed and loaded
  packages <- c(
    "labelled",
    "rstatix",
    "ggpubr",
    "GGally",
    "car",
    "Epi",
    "lme4",
    "lmerTest",
    "emmeans",
    "multcomp",
    "geepack",
    "ggeffects",
    "gt",
    "readxl",
    "data.table",
    "tidyverse",
    "summarytools",
    "ggplot2",
    "dplyr",
    "lubridate",
    "readr",
    "stringr",
    "tibble",
    "naniar",
    "NHANES",
    "forcats",
    "moments"
  )
  
  # Install packages that are not already installed
  installed_packages <- installed.packages()
  for (pkg in packages) {
    if (!(pkg %in% installed_packages[, "Package"])) {
      install.packages(pkg)
    }
  }
  
  # Load all the packages
  lapply(packages, library, character.only = TRUE)
}

# Run the function to install and load all packages
install_and_load_packages()

#Relevant libraries for use in this data 
library(labelled)   # This is used to labeling data
library(rstatix)    # This is used for summary statistics
library(ggpubr)     # This is used for convenient summary statistics and plots
library(GGally)     # This is used for advanced plot
library(car)        # This is used for useful for anova/wald test
library(Epi)        # This is used for easy getting CI for model coef/pred
library(lme4)       # This is used for linear mixed-effects models
library(lmerTest)   # This is used for test for linear mixed-effects models
library(emmeans)    # This is used for marginal means
library(multcomp)   # This is used for CI for linear combinations of model coef
library(geepack)    # This is used for generalized estimating equations
library(ggeffects)  # This is used for marginal effects, adjusted predictions
library(gt)         # This is used for nice tables
library(readxl)     # This is used for reading Excel files
library(data.table) # This is used for fast data manipulation
library(tidyverse)  # for data manipulation and visualization
library(summarytools) # This is used for detailed summary statistics
library(ggplot2)    # This is used for data visualization
library(dplyr)      #This is used  for data manipulation
library(lubridate)  # This is used for date-time manipulation
library(readr)      # This is used for reading data
library(stringr)    # This is used for string manipulation
library(tibble)     # This is used for modern data frames
library(naniar)     # This is used for handling missing data
library(NHANES)     # This is used for accessing NHANES data
library(forcats)    # This is used for working with categorical variables
library(moments)
```

# Step 1
# This is the importation of the longitudinal data and turning it into a dataframe 
```{r}
# DATA IMPORTATION 
longi_data <- read.csv("C:/Users/aoogbomo/OneDrive - University of Bradford/Desktop/mimiproject/ageing_better_cmf.csv", header = TRUE, stringsAsFactors = FALSE)
head(longi_data)
```
# Check the summary and the structure of the data
```{r}
summary(longi_data)
str(longi_data)
```

# Convert all the -1 to NA within the large Data frame
```{r}
# To work with this data we first convert all the -1 to NA as specified by the data dictionary 
longi_new_data<- longi_data %>%
  mutate(across(everything(), ~ {
    ifelse(. == -1 | . == "-1", NA, .)
  }))
longi_new_data
glimpse(longi_new_data)
```
# looping the longi_new_data
```{r}
for (col_name in colnames(longi_new_data)) {
  # Create a bar chart for the current column
  p <- ggplot(longi_new_data, aes_string(x = col_name)) +
    geom_bar() +
    ggtitle(paste("Bar Chart of", col_name)) +
    xlab(col_name) +
    ylab("Count")
  
  # Print the plot
  print(p)
}
```
# Remove columns with 100% missing values or NA 
```{r}
# Remove columns with 100% NA values
longi_new_data_clean <- longi_new_data %>%
  select(where(~ !all(is.na(.))))

# View the cleaned dataframe
print(head(longi_new_data_clean))

# Check the columns that were removed
removed_columns <- setdiff(names(longi_new_data), names(longi_new_data_clean))
cat("Removed columns with 100% NA values:\n")
print(removed_columns)
```
```{r}
variable_classes <- sapply(longi_new_data_clean, class)
print(variable_classes)
```


# Step 2. The first partition will be the Response variables
```{r}
# We set what we know as the Response variables and sebset them from the main data
response_columns<- c("DEJONGscore", "DEJONG_FUp1_Score", "DEJONG_FUp2_Score", 
                       "DEJONG_FUp3_Score", "DEJONG_FUp4_Score", "DEJONG_FUp5_Score",
                       "UCLA1Average", "UCLA_Fup1_Average", "UCLA_FUp2_Average", 
                       "UCLA_FUp3_Average", "UCLA_FUp4_Average", "UCLA_FUp5_Average",
                       "EQVASScore", "EQVAS_FUp1_score", "EQVAS_FUp2_score", 
                       "EQVAS_FUp3_score", "EQVAS_FUp4_score", "EQVAS_FUp5_score",
                       "SWEMWBSScore", "SWEMWBS_FUp1_score_num", "SWEMWBS_FUp2_score_num", 
                       "SWEMWBS_FUp3_score_num", "SWEMWBS_FUp4_score_num", "SWEMWBS_FUp5_score_num")

# Subset the data frame
Response <- longi_new_data_clean %>% select(all_of(response_columns))

# View the subsetted data frame
print(Response)
```

# what are the classes present in the response
```{r}
classes_Response <- sapply(Response, class)
print(classes_Response)
```


# This is the  exploratory data analysis for the response variables here we first draw a barplot to visualise the response variables 
```{r}
for (col_name in colnames(Response)) {
  # Create a bar chart for the current column
  p <- ggplot(Response, aes_string(x = col_name)) +
    geom_bar() +
    ggtitle(paste("Bar Chart of", col_name)) +
    xlab(col_name) +
    ylab("Count")
  
  # Print the plot
  print(p)
}
```
# we proceed with a histogram and divide the response into bins, because from the above we recognise that this is a numerical variable.  
```{r}
# Loop through each column in the data frame
# Load necessary library
library(ggplot2)

# Loop through each column in the data frame
for (col_name in colnames(Response)) {
  
  # Convert the column to numeric if it's not already
  numeric_column <- as.numeric(as.character(Response[[col_name]]))
  
  # Handle any NA values generated during conversion
  numeric_column <- numeric_column[!is.na(numeric_column)]
  
  # Create a histogram for the numeric data, grouped into bins
  p <- ggplot(data.frame(numeric_column), aes(x = numeric_column)) +
    geom_histogram(binwidth = diff(range(numeric_column)) / 30, fill = "blue", color = "black") +
    ggtitle(paste("Histogram of", col_name, " (binned)")) +
    xlab(col_name) +
    ylab("Frequency")
  
  # Print the plot
  print(p)
}
```
# The next step is to test for normality , Skewness,Kurtosis
```{r}
# Loop through each column in the data frame
for (col_name in colnames(Response)) {
  
  # Convert the column to numeric if it's not already
  numeric_column <- as.numeric(as.character(Response[[col_name]]))
  
  # Remove NA values
  numeric_column <- numeric_column[!is.na(numeric_column)]
  
  # Skip the test if the column has fewer than 3 unique values
  if(length(unique(numeric_column)) < 3) {
    cat("Skipping", col_name, "- not enough unique values.\n")
    next
  }
  
  cat("\nTesting column:", col_name, "\n")
  
  # Summary Statistics
  cat("Summary Statistics:\n")
  print(summary(numeric_column))
  cat("Skewness:", skewness(numeric_column), "\n")
  cat("Kurtosis:", kurtosis(numeric_column), "\n")
  
  # Shapiro-Wilk Test for Normality
  if (length(numeric_column) >= 3 & length(numeric_column) <= 5000) {
    cat("Shapiro-Wilk Normality Test:\n")
    tryCatch({
      print(shapiro.test(numeric_column))
    }, error = function(e) {
      cat("Error in Shapiro-Wilk Test:", e$message, "\n")
    })
  } else {
    cat("Skipping Shapiro-Wilk Test: Sample size not between 3 and 5000.\n")
  }
  
  # Q-Q Plot
  qqnorm(numeric_column, main = paste("Q-Q Plot of", col_name))
  qqline(numeric_column, col = "red")
  
  # Breusch-Pagan Test for Homoscedasticity
  # First, fit a linear model (assuming the dependent variable is numeric_column)
  lm_model <- lm(numeric_column ~ 1)
  cat("Breusch-Pagan Test for Homoscedasticity:\n")
  tryCatch({
    print(bptest(lm_model))
  }, error = function(e) {
    cat("Error in Breusch-Pagan Test:", e$message, "\n")
  })
  
  cat("\n----------------------------\n")
}

```

# We go futher to explore the response by the Gender 
```{r}
# we first convert to the long format.
longi_new_data_long <- longi_new_data_clean %>%
  pivot_longer(cols = all_of(response_columns), 
               names_to = "Response_Variable", 
               values_to = "Value")

# Loop through each column in the response columns and plot
for (col_name in response_columns) {
  # Filter the data for the current response variable
  plot_data <- longi_new_data_long %>% filter(Response_Variable == col_name)
  
  # Create a bar chart colored by Gender
  p <- ggplot(plot_data, aes(x = Response_Variable, y = Value, fill = interaction(Gender))) +
    geom_bar(stat = "identity", position = "dodge") +
    ggtitle(paste("Bar Chart of", col_name, "coloured by Gender")) +
    xlab(col_name) +
    ylab("Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Print the plot
  print(p)
}
```

```{r}
# Loop through each response variable column and plot
for (col_name in response_columns) {
  # Filter the data for the current response variable
  plot_data <- longi_new_data_long %>% filter(Response_Variable == col_name)
  
  # Create a bar chart colored by Gender
  p <- ggplot(plot_data, aes(x = Value, fill = Gender)) +
    geom_bar(position = "dodge") +
    ggtitle(paste("Bar Chart of", col_name, "by Gender")) +
    xlab(col_name) +
    ylab("Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Print the plot
  print(p)
}
```


# We explore the interaction of response varable with the birth year 
```{r}
# Loop through each column in the response columns and plot
for (col_name in response_columns) {
  # Filter the data for the current response variable
  plot_data <- longi_new_data_long %>% filter(Response_Variable == col_name)
  
  # Create a bar chart colored by Birthyear_7Categories
  p <- ggplot(plot_data, aes(x = Response_Variable, y = Value, fill = interaction(Birthyear_7Categories))) +
    geom_bar(stat = "identity", position = "dodge") +
    ggtitle(paste("Bar Chart of", col_name, "Birthyear_7Categories")) +
    xlab(col_name) +
    ylab("Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Print the plot
  print(p)
}
```

# We explore the interaction of response varable with the Ethnicity

```{r}
# Loop through each column in the response columns and plot the interaction with the Ethnic_5Categories
for (col_name in response_columns) {
  # Filter the data for the current response variable
  plot_data <- longi_new_data_long %>% filter(Response_Variable == col_name)
  
  # Create a bar chart colored by Ethnic_5Categories
  p <- ggplot(plot_data, aes(x = Response_Variable, y = Value, fill = interaction(Ethnic_5Categories))) +
    geom_bar(stat = "identity", position = "dodge") +
    ggtitle(paste("Bar Chart of", col_name, "Ethnicity")) +
    xlab(col_name) +
    ylab("Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Print the plot
  print(p)
} 
```

# We explore the interaction of response varable with  Sexuality
```{r}
# Loop through each column in the response columns and plot
for (col_name in response_columns) {
  # Filter the data for the current response variable
  plot_data <- longi_new_data_long %>% filter(Response_Variable == col_name)
  
  # Create a bar chart colored by Sexuality
  p <- ggplot(plot_data, aes(x = Response_Variable, y = Value, fill = interaction(Sexuality_2Categories))) +
    geom_bar(stat = "identity", position = "dodge") +
    ggtitle(paste("Bar Chart of", col_name, "Birthyear_7Categories")) +
    xlab(col_name) +
    ylab("Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Print the plot
  print(p)
}
```


```{r}
# Thats the first part of the exploratory data analysis between response variable and some of the demographic,#
# next  o the next part which is manipulating the dataset 
```


# Step 3 is the Second partition will be the  activity_start_enddate 
```{r}
month_start_end_cols <- grep("MonthStart|MonthEnd", names(longi_new_data_clean), value = TRUE)
# Subset the data frame to include only these columns
activity_start_enddate <- longi_new_data_clean[, month_start_end_cols]
print(activity_start_enddate)
```

```{r}
activity_start_enddate
```


```{r}
start_end_columns=colnames(activity_start_enddate)
```

```{r}
str(activity_start_enddate)
```

```{r}
summary(activity_start_enddate)
```

```{r}
# Loop through each column and count the values
for (col_name in colnames(activity_start_enddate)) {
  cat("Counts for column:", col_name, "\n")
  print(table(activity_start_enddate[[col_name]]))
  cat("\n")
}
```

```{r}
# Using lapply to check unique values for each column
unique_values <- lapply(activity_start_enddate, unique)

# Print unique values for each column
for (col_name in names(unique_values)) {
  cat("Unique values for column:", col_name, "\n")
  print(unique_values[[col_name]])
  cat("\n")
}
activity_start_enddate %>%
  gather(key = "col_variable", value = "value") %>%
  group_by(col_variable) %>%
  summarise(unique_values = n_distinct(value)) %>%
  arrange(col_variable)
```


```{r}
glimpse(unique_values)
```


```{r}
# 
# Initialize a new dataframe to store the differences
difference_in_days <- data.frame(matrix(ncol = 0, nrow = nrow(activity_start_enddate)))

# Loop through the columns to identify pairs with MonthStart and MonthEnd
for (col_name in names(activity_start_enddate)) {
  if (grepl("MonthStart$", col_name)) {
    prefix <- sub("MonthStart$", "", col_name)
    start_col <- paste0(prefix, "MonthStart")
    end_col <- paste0(prefix, "MonthEnd")
    
    if (start_col %in% names(activity_start_enddate) && end_col %in% names(activity_start_enddate)) {
      # Convert date strings to Date format
      start_dates <- ymd(paste0(activity_start_enddate[[start_col]], "-01"), quiet = TRUE)
      end_dates <- ymd(paste0(activity_start_enddate[[end_col]], "-01"), quiet = TRUE)
      
      # Calculate the difference in months, ensuring NA is returned if any date is NA
      date_diff <- ifelse(is.na(start_dates) | is.na(end_dates), NA, interval(start_dates, end_dates) %/% months(1))
      
      # Create a new column name for the difference with _diff suffix
      diff_col_name <- paste0(sub("_MonthStart$", "", start_col), "_diff")
      
      # Add the difference to the new dataframe
      difference_in_days[[diff_col_name]] <- date_diff
    }
  }
}

# Print the new dataframe with the differences
print(difference_in_days)
```






```{r}
# Using lapply to check unique values for each column
unique_value_dif <- lapply(difference_in_days, unique)

# Print unique values for each column
for (col_name in names(unique_value_dif)) {
  cat("Unique values for column:", col_name, "\n")
  print(unique_value_dif[[col_name]])
  cat("\n")
}
activity_start_enddate %>%
  gather(key = "col_variable", value = "value") %>%
  group_by(col_variable) %>%
  summarise(unique_value_dif = n_distinct(value)) %>%
  arrange(col_variable)
```

```{r}
#EDA to explore the plot of the difference in columns. 
# Clean column names
colnames(difference_in_days) <- make.names(colnames(difference_in_days))
# Loop through each column and plot a bar chart
for (col_name in colnames(difference_in_days)) {
  # Convert column to factor if it's not already
  if (!is.factor(difference_in_days[[col_name]])) {
   difference_in_days[[col_name]] <- as.factor(difference_in_days[[col_name]])
  }
  
# Create a bar chart for the current column
  p <- ggplot(difference_in_days, aes_string(x = col_name)) +
    geom_bar() +
    ggtitle(paste("Bar Chart of", col_name)) +
    xlab(col_name) +
    ylab("Count") +
    theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1)) +  # Rotate x-axis text
    theme(plot.margin = margin(10, 10, 10, 100)) +  # Add larger margin to avoid clipping
    theme(axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)))  # Increase x-axis title margin
  
  # Print the plot
  print(p)
}
```

```{r}
colnames(difference_in_days) <- make.names(colnames(difference_in_days))

# Loop through each column and plot a histogram
for (col_name in colnames(difference_in_days)) {
  # Convert column to numeric if it's not already
  if (!is.numeric(difference_in_days[[col_name]])) {
    difference_in_days[[col_name]] <- as.numeric(difference_in_days[[col_name]])
  }
  
  # Create a histogram for the current column
  p <- ggplot(difference_in_days, aes_string(x = col_name)) +
    geom_histogram(binwidth = 1, fill = "blue", color = "black") +
    ggtitle(paste("Histogram of", col_name)) +
    xlab(col_name) +
    ylab("Frequency") +
    theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1)) +  # Rotate x-axis text
    theme(plot.margin = margin(10, 10, 10, 10))  # Add larger margin to avoid clipping
  
  # Print the plot
  print(p)
}

```


```{r}
# Initialize a new dataframe to store the 1/0 values
binary_df <- data.frame(matrix(ncol = 0, nrow = nrow(activity_start_enddate)))

# Loop through the columns to identify pairs with MonthStart and MonthEnd
for (col_name in names(activity_start_enddate)) {
  if (grepl("MonthStart$", col_name)) {
    prefix <- sub("MonthStart$", "", col_name)
    start_col <- paste0(prefix, "MonthStart")
    end_col <- paste0(prefix, "MonthEnd")
    
    if (start_col %in% names(activity_start_enddate) && end_col %in% names(activity_start_enddate)) {
      # Create a new column with 1 if both MonthStart and MonthEnd have date values, 0 if MonthStart is NA
      presence_col <- ifelse(!is.na(activity_start_enddate[[start_col]]) & !is.na(activity_start_enddate[[end_col]]), 1, 0)
      
      # Create a new column name for the presence check with _presence suffix
      presence_col_name <- paste0(sub("_MonthStart$", "", start_col), "_binary")
      
      # Add the presence column to the new dataframe
      binary_df[[presence_col_name]] <- presence_col
    }
  }
}

# Print the new dataframe with the presence columns
print(binary_df) 
```

```{r}
# Using lapply to check unique values for each column
unique_value_bin <- lapply(binary_df, unique)

# Print unique values for each column
for (col_name in names(unique_value_bin)) {
  cat("Unique values for column:", col_name, "\n")
  print(unique_value_bin[[col_name]])
  cat("\n")
}
activity_start_enddate %>%
  gather(key = "col_variable", value = "value") %>%
  group_by(col_variable) %>%
  summarise(unique_value_bin = n_distinct(value)) %>%
  arrange(col_variable)
```


```{r}
#EDA to explore the plot of the difference in columns. 
# Clean column names
colnames(binary_df) <- make.names(colnames(binary_df))
# Loop through each column and plot a bar chart
for (col_name in colnames(binary_df)) {
  # Convert column to factor if it's not already
  if (!is.factor(binary_df[[col_name]])) {
    binary_df[[col_name]] <- as.factor(binary_df[[col_name]])
  }
  
# Create a bar chart for the current column
  p <- ggplot(binary_df, aes_string(x = col_name)) +
    geom_bar() +
    ggtitle(paste("Bar Chart of", col_name)) +
    xlab(col_name) +
    ylab("Count") +
    theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1)) +  # Rotate x-axis text
    theme(plot.margin = margin(10, 10, 10, 100)) +  # Add larger margin to avoid clipping
    theme(axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)))  # Increase x-axis title margin
  
  # Print the plot
  print(p)
}
```




#Step 4; 
# The other columns in the dataframe are yet to be class as interventions or cofounding factors or droped out after proper exploration so we name it as longi_new_data_clean_r

# Remove the 'response' column
```{r}
longi_new_data_clean_r<- longi_new_data_clean %>% select(-response_columns)
longi_new_data_clean_r
```

# validate that the response columns are gone from the longi_new_data_clean

```{r}
columns_to_check <- c("DEJONGscore", "DEJONG_FUp1_Score", "DEJONG_FUp2_Score", 
                      "DEJONG_FUp3_Score", "DEJONG_FUp4_Score", "DEJONG_FUp5_Score",
                      "UCLA1Average", "UCLA_Fup1_Average", "UCLA_FUp2_Average", 
                      "UCLA_FUp3_Average", "UCLA_FUp4_Average", "UCLA_FUp5_Average",
                      "EQVASScore", "EQVAS_FUp1_score", "EQVAS_FUp2_score", 
                      "EQVAS_FUp3_score", "EQVAS_FUp4_score", "EQVAS_FUp5_score",
                      "SWEMWBSScore", "SWEMWBS_FUp1_score_num", "SWEMWBS_FUp2_score_num", 
                      "SWEMWBS_FUp3_score_num", "SWEMWBS_FUp4_score_num", "SWEMWBS_FUp5_score_num")

# Check which columns are present in the longi_new_data_clean_r dataframe
present_columns <- columns_to_check[columns_to_check %in% names(longi_new_data_clean_r)]

# Print the result
if (length(present_columns) > 0) {
  cat("The following columns from Response dataframe are present in longi_new_data_clean_r:\n")
  print(present_columns)
} else {
  cat("None of the specified columns from Response dataframe are present in longi_new_data_clean_r.\n")
}
```

# Now we remove the activity start and end date from the new dataframe to arrive at a column without both response, ens and start date. 
```{r}
start_end_columns=colnames(activity_start_enddate)
longi_new_data_clean_r_e <-longi_new_data_clean_r %>% select(-start_end_columns)
longi_new_data_clean_r_e
```


```{r}
# Check which columns are present in the longi_new_data_clean_r dataframe
present_date_columns <- start_end_columns[start_end_columns %in% names(longi_new_data_clean_r_e)]

# Print the result
if (length(present_date_columns) > 0) {
  cat("The following columns from activity_start_enddate dataframe are present in longi_new_data_clean_r_e:\n")
  print(present_date_columns)
} else {
  cat("None of the specified columns from activity_start_enddate dataframe are present in longi_new_data_clean_r_e.\n")
}
```

```{r}
 longi_new_data_clean_r_e

```

#############################################################################
```{r}
# Clean column names
colnames(longi_new_data_clean_r_e) <- make.names(colnames(longi_new_data_clean_r_e))
# Loop through each column and plot a bar chart
for (col_name in colnames(longi_new_data_clean_r_e)) {
  # Convert column to factor if it's not already
  if (!is.factor(longi_new_data_clean_r_e[[col_name]])) {
    longi_new_data_clean_r_e[[col_name]] <- as.factor(longi_new_data_clean_r_e[[col_name]])
  }
  #############
  # Create a bar chart for the current column
  p <- ggplot(longi_new_data_clean_r_e, aes_string(x = col_name)) +
    geom_bar() +
    ggtitle(paste("Bar Chart of", col_name)) +
    xlab(col_name) +
    ylab("Count")
  
  # Print the plot
  print(p)
}
```
# here we have our dataframe broken down into four dataframe from where we will do another EDA to make some determinations. 
```{r}

# we concatenate the dataframe
# Data without start and end month
longi_new_data_clean_r_e
# Data with only the duration of treatment in days 
difference_in_days
binary_df
Response 
```

# We concatenate this dataframes are named Response, difference_in_days, binary_df, and class_age_data into a new big dataframe for our analysis
```{r}
# Concatenate the dataframes by columns
combined_longi <- cbind(Response, difference_in_days, binary_df, longi_new_data_clean_r_e)

# Display the concatenated dataframe
print(combined_longi)

```




# Firstly we convert the date of birth to years in the new wangeled data


```{r}
# first we handle the "Birthyear_7Categories" column by finding the categories, this is to enable us determine the age range
unique_Birthyear_7Categories <- unique(combined_longi$Birthyear_7Categories)

print(unique_Birthyear_7Categories)
```


```{r}
#By using this categories ("1920-1929"      "1930-1939"      "1940-1949"      "1950-1959"      "1960-1969"      NA               "1970 and after" "Before 1920") we will create a category that we can use to mutate the combined_longi to create a new dataframe longi_new_data


# Define the replacement function
replace_birthyear_categories <- function(category) {
  case_when(
    category == "1920-1929" ~ "100-91",
    category == "1930-1939" ~ "90-81",
    category == "1940-1949" ~ "80-71",
    category == "1950-1959" ~ "70-61",
    category == "1960-1969" ~ "60-51",
    is.na(category) ~ NA_character_,
    category == "1970 and after" ~ "50-0",
    category == "Before 1920" ~ ">100",
    TRUE ~ category  # 
  )
}
# Apply the replacement function to the column
longi_new_data <- combined_longi %>%
mutate(Age_Range = replace_birthyear_categories(Birthyear_7Categories))
```

```{r}
#drop the "Birthyear_7Categories" column and bring the Age_Range forward to the second position. 
# to acheive this we make use of the replacement function
longi_new_data  <- longi_new_data  %>%
  mutate(Age_Range = replace_birthyear_categories(Birthyear_7Categories)) %>%
  select(-Birthyear_7Categories) %>%  # Drop the Birthyear_7Categories column
  select(1, Age_Range, everything())  # Move Age_Range to the second position

# Print the updated data frame
print(longi_new_data)
```
  
  

  
 



# The next stage will know the response  and the duration
```{r}
# We set what we know as the Response variables and sebset them from the main data
mean_response_columns<- c("DEJONGscore","UCLA1Average","EQVASScore","SWEMWBSScore")

# Subset the data frame
mean_response_columns<- longi_new_data_clean %>% select(all_of(mean_response_columns))

# View the subsetted data frame
print(mean_response_columns)
```



```{r}  
str(Response)
# Get the data types of each column
data_types <- sapply(Response, class)
# Print the data types
print(data_types)
```


 




# Reshape Data for longitudinal analysis 


```{r}
# Reshape data from wide to long format
Response_long_data <- Response %>%
  pivot_longer(
    cols = everything(), # Convert all columns
    names_to = "Measurement",
    values_to = "Score"
  ) %>%
  mutate(
    Year = case_when(
      str_detect(Measurement, "FUp1|Fup1") ~ "Year 1",
      str_detect(Measurement, "FUp2|Fup2") ~ "Year 2",
      str_detect(Measurement, "FUp3|Fup3") ~ "Year 3",
      str_detect(Measurement, "FUp4|Fup4") ~ "Year 4",
      str_detect(Measurement, "FUp5|Fup5") ~ "Year 5",
      TRUE ~ "Baseline"  # Default to Baseline if no FUp pattern matched
    )
  )

# View reshaped data
print(Response_long_data)
```


```{r}
group_by(Response_long_data, Year) %>% 
  get_summary_stats(Score)
```

# 

```{r}
# make a box plot for the distribution of the response 
p <- ggplot(Response_long_data, aes(x = Measurement, y = Score, fill = Measurement)) +
  geom_boxplot(width = 0.75) +  # Increase the width of the box plots to make them more prominent
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +  # Enhance jitter visibility
  theme_minimal() +
  guides(fill = FALSE) +  # Turn off the legend for fill
  labs(x = "", y = "Score", title = "Distribution of Scores by Measurement") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),  # Rotate and adjust text size
    plot.title = element_text(size = 16, hjust = 0.5),  # Center and enlarge the title
    axis.title = element_text(size = 14)  # Enlarge axis titles
  )

# Print the plot
print(p)

```

```{r}
# Create a new column for the Metric name
Response_long_data <- Response_long_data %>%
  mutate(Metric = ifelse(grepl("FUp", Measurement), 
                         sub("_.*", "", Measurement), 
                         Measurement))

# Calculate the mean score for each year
mean_scores <- Response_long_data %>%
  group_by(Year) %>%
  summarise(MeanScore = mean(Score, na.rm = TRUE)) %>%
  arrange(factor(Year, levels = c('Baseline', 'Year 1', 'Year 2', 'Year 3', 'Year 4', 'Year 5')))

ggplot(mean_scores, aes(x = Year, y = MeanScore, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean Response Over Time", x = "Year", y = "Mean Score") +
  theme_minimal()

```

```{r}


# Group by Year and Metric, then calculate the mean and confidence intervals
mean_scores <- Response_long_data %>%
  filter(grepl("FUp", Measurement)) %>%  # Filter only the FUp measurements
  group_by(Year, Metric) %>%  # Group by Year and Metric (extracted earlier)
  summarise(mean = mean(Score, na.rm = TRUE),
            ymin = mean(Score, na.rm = TRUE) - 1.96 * sd(Score, na.rm = TRUE) / sqrt(n()),
            ymax = mean(Score, na.rm = TRUE) + 1.96 * sd(Score, na.rm = TRUE) / sqrt(n()), 
            .groups = "drop") %>%
  ungroup() %>%
  mutate(Year_numeric = as.numeric(gsub("Year ", "", Year)) - 0.05 + 0.05 * (Metric == "Boy"))  # Adjust for year visualization

# Plot the data
ggplot(mean_scores, aes(x = Year_numeric, y = mean, col = Metric, shape = Metric)) +
  geom_point() +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.2) +
  geom_line() +
  labs(title = "Response Trajectories over time", x = "Year", y = "Mean Score", shape = "Metric", col = "Metric") +
  theme_minimal()
```


```{r}
# Filter the data to include only follow-up measurements
fup_data <- Response_long_data %>%
  filter(grepl("FUp", Measurement))

# Extract the year number for grouping
fup_data <- fup_data %>%
  mutate(YearNumber = factor(sub(".*FUp([1-5]).*", "\\1", Measurement)))

# Calculate the mean score for each year
mean_scores <- fup_data %>%
  group_by(YearNumber) %>%
  summarise(MeanScore = mean(Score, na.rm = TRUE))

# Plot the mean scores for each follow-up year
ggplot(mean_scores, aes(x = YearNumber, y = MeanScore)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean Scores for Each Follow-Up Year", x = "Year", y = "Mean Score") +
  theme_minimal()
```

```{r}
# Reshape the data to long format
Response_long_data <- Response %>%
  pivot_longer(cols = starts_with("DEJONG") | starts_with("UCLA") | starts_with("EQVAS") | starts_with("SWEMWBS"), 
               names_to = "Measurement", values_to = "Score") %>%
  mutate(YearNumeric = case_when(
    grepl("FUp1", Measurement) ~ 1,
    grepl("FUp2", Measurement) ~ 2,
    grepl("FUp3", Measurement) ~ 3,
    grepl("FUp4", Measurement) ~ 4,
    grepl("FUp5", Measurement) ~ 5,
    TRUE ~ 0  # This is for the baseline or first year measurements like DEJONGscore, UCLA1Average, etc.
  ))

# Filter out the baseline if you want to include only follow-up years
filtered_data <- Response_long_data %>%
  filter(YearNumeric > 0) %>%  # Keep only the follow-up data if required
  drop_na(Score)  # Drop rows where Score is NA

# Plot the data
ggplot(filtered_data, aes(x = YearNumeric, y = Score, col = Measurement)) +
  geom_point(na.rm = TRUE) +
  geom_line(na.rm = TRUE) +
  facet_wrap(~ Measurement, scales = "free_y") +  # Facet by Measurement
  labs(x = "Year", y = "Score", col = "Measurement") +
  guides(col = guide_legend(nrow = 3)) +
  theme_minimal()
 
```

```{r}



Response_long_data <- Response_long_data %>%
  mutate(YearNumeric = case_when(
    grepl("FUp1", Measurement) ~ 1,
    grepl("FUp2", Measurement) ~ 2,
    grepl("FUp3", Measurement) ~ 3,
    grepl("FUp4", Measurement) ~ 4,
    grepl("FUp5", Measurement) ~ 5,
    TRUE ~ NA_real_  # Exclude baseline scores or handle them separately
  ))

# Filter out rows that don't match the follow-up pattern
filtered_data <- Response_long_data %>%
  filter(!is.na(YearNumeric))  # Keep only the follow-up data

# Create the scatter plot




# Replace 'Metric' with the correct column name, likely 'Measurement'
ggplot(filtered_data, aes(x = Measurement, y = Score, col = Measurement)) +
  geom_point(size = 3) +  # Scatter plot with points
  facet_wrap(~ YearNumeric) +  # Facet by YearNumeric
  labs(x = "Measurement", y = "Score", col = "Measurement") +
  theme_minimal()


```

```{r}

# Reshape the data to long format
Response_long_data <- Response %>%
  pivot_longer(
    cols = starts_with("DEJONG") | starts_with("UCLA") | starts_with("EQVAS") | starts_with("SWEMWBS"), 
    names_to = "Measurement", values_to = "Score"
  ) %>%
  mutate(
    YearNumeric = case_when(
      grepl("FUp1", Measurement) ~ 1,
      grepl("FUp2", Measurement) ~ 2,
      grepl("FUp3", Measurement) ~ 3,
      grepl("FUp4", Measurement) ~ 4,
      grepl("FUp5", Measurement) ~ 5,
      TRUE ~ 0  # This is for the baseline or first year measurements like DEJONGscore, UCLA1Average 
    )
  )

# Filter out the baseline if you want to include only follow-up years
filtered_data <- Response_long_data %>%
  filter(YearNumeric > 0) %>%  # Keep only the follow-up data if required
  drop_na(Score)  # Drop rows where Score is NA

# Plot the data
ggplot(filtered_data, aes(x = YearNumeric, y = Score, col = factor(Measurement))) +
  geom_line() +
  labs(x = "Year", y = "Score", col = "Measurement") +
  guides(col = guide_legend(nrow = 3)) +
  theme_minimal()

```


# Combined data is combined_longi
```{r} 
# longi_new_data is the result of the concatenation
longi_new_data
```

# Analysis 
```{r} 
#library(tidyr)
#
  # The first thing is to convert all the column for transformation to numerical variable 
# Convert all relevant columns to numeric
finalda <- longi_new_data %>%
  mutate(across(starts_with("DEJONG"), as.numeric),
         across(starts_with("UCLA"), as.numeric),
         across(starts_with("EQVAS"), as.numeric),
         across(starts_with("SWEMWBS"), as.numeric))
```

```{r}
# Convert all relevant columns to numeric
finalda <- longi_new_data %>%
  mutate(across(starts_with("DEJONG"), as.numeric),
         across(starts_with("UCLA"), as.numeric),
         across(starts_with("EQVAS"), as.numeric),
         across(starts_with("SWEMWBS"), as.numeric))

# 
long_data <- finalda %>%
  pivot_longer(cols = starts_with("DEJONG") | starts_with("UCLA") | starts_with("EQVAS") | starts_with("SWEMWBS"),
               names_to = c("Measure", "Time"),
               names_pattern = "(.*)_(FUp[0-5])",
               values_to = "Value") %>%
  mutate(Time = recode(Time, "FUp0" = "Baseline", "FUp1" = "Year1", "FUp2" = "Year2", "FUp3" = "Year3", "FUp4" = "Year4", "FUp5" = "Year5"))

# View the transformed data
head(long_data)
```
```{r} 
str(long_data)
```
```{r} 
colnames(long_data)
```



```{r} 

# Univariate EDA - Histogram
#library(ggplot2)

ggplot(long_data, aes(x = Value)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  facet_wrap(~ Measure, scales = "free_x") +
  theme_minimal() +
  labs(title = "Distribution of Values for Each Measure")

```



```{r} 
# Bivariate EDA - Scatter plot
ggplot(long_data, aes(x = Time, y = Value, color = Gender)) +
  geom_point() +
  facet_wrap(~ Measure) +
  theme_minimal() +
  labs(title = "Bivariate Analysis: Value over Time for Each Measure",
       x = "Time", y = "Value")


```



```{r} 
summary(long_data)
long_data <- na.omit(long_data) # Simple removal of rows with NAs
```

```{r} 
#Outlier Detection:

#Identify and address outliers which could skew the analysis.


boxplot(long_data$Value ~ long_data$Measure)

```


```{r} 
#Correlation Analysis:

#This will be explored after the removal of NA
#print(cor_matrix)
```


```{r} 
#Interaction Effects:
#interaction.plot(long_data$Time, long_data$Measure, long_data$Value)
```



```{r} 
#Dimensionality Reduction

```


```{r} 
library(lme4)

# Initialize lists to store the results
model_summaries <- list()
anova_results <- list()

# Loop through each column in longi_new_data_clean_r_e to use as predictors
for (col_name in colnames(longi_new_data_clean_r_e)) {
  
  # Create a formula dynamically using Measure as the response variable
  formula <- as.formula(paste("Measure ~", col_name, "+ (1 | Ethnic_5Categories)"))
  
  # Try fitting the model and catching any errors
  try({
    # Fit the mixed-effects model
    model <- lmer(formula, data = long_data)
    
    # Store the summary of the model
    model_summaries[[col_name]] <- summary(model)
    
    # Store the ANOVA results
    anova_results[[col_name]] <- anova(model)
    
    # Print the summary for each model
    print(paste("Summary for model with predictor:", col_name))
    print(summary(model))
    
    # Print the ANOVA results for each model
    print(paste("ANOVA for model with predictor:", col_name))
    print(anova(model))
  }, silent = TRUE) # This will prevent the loop from stopping if an error occurs
}

#save(model_summaries, anova_results, file = "model_results.RData")

```
